Adaptive evolution with aneuploidy and mutation

### 2022-01-12 ms/ms.tex
revised parameter inference section (w->W as w is fitness, ref, phrasing)

\paragraph{Likelihood function.} 
Because our model, just like the Wright-Fisher model, is non-linear and stochastic, computing the distribution of fixation time $T(g)$ of genotype $g$ for use in the likelihood function is intractable (it is even hard to use a diffusion-equation approximation due to the model having multiple genotypes, rather than just two).
We overcome this problem by approximating the likelihood using simulations. We simulate 1,000 experiments per parameter vector $\theta = (\mu, \delta, s, b, c)$, resulting in a set of simulated observations $\tilde{\vec X} = \{\tilde{X}_i\}_{i=1}^{1000}$. We then compute the approximate likelihood,
\begin{equation}\begin{aligned}
\label{eq:heatstress-likelihood}
\likelihood(\theta) = &\ P^4(200 \le T(2n+1) \le 450) \cdot 
	\Big[1 - \\
	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)- \\
	&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big)+ \\
	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big) 
	\Big]\;,
\end{aligned}\end{equation}
where $!\{\ldots\}$ is the "logical not" operator, $P^4(\ldots)$ is the 4th power of $P(\ldots)$, and all probabilities $P_{\tilde{\vec X}}(\ldots)$ are approximated from the results of the simulations $\tilde{\vec X}$. For example, $P_{\tilde{\vec X}}\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)$ is approximated by taking simulations in which \anwt\; fixed before generation 450 but not before generation 200, and computing the fraction of such simulations in which \eumt\; did not fix by generation 1,700, and hence aneuploidy did not extinct before generation 1,700.

@@

% TODO if we show results of multi-locus model, write how much replicates were used.
% TODO ref to place where we show 1,000 is enough

@@

%% YR revise
The ABS-SMC algorithm employs sequential importance sampling over multiple iterations~\citep{Toni2009, Klinger2017, Syga2021}.
In iteration $t$ of the algorithm, a set of parameter vectors, $\{\theta_{i,t}\}_{i=1}^{n_t}$, also called \emph{particles}, are constructed in the following way.
A proposal particle, $\theta^*$, is sampled from a proposal distribution, and is either accepted or rejected, until $n_t$ particles are accepted.
The number of particles, $n_t$, is adapted at every iteration $t$ using the adaptive population strategy.
For $t=0$, the proposal particle is sampled from the prior distribution, $p(\theta)$.
For $t>0$, the proposal particle is sampled from the particles accepted in the previous iteration, $\{\theta_{i,t-1}\}_{i=1}^{n_{t-1}}$, each with a probability relative to its weight $w_{t-1}(\theta_{i,t-1})$ (see below). The proposal particle is then perturbed using a kernel perturbation kernel, $K_t(\theta^* \mid \theta)$ where $\theta$ is the sample from the previous iteration.
Then, a set of synthetic observations $\tilde{\vec X}^*$ is simulated, and the proposal particle $\theta^*$ is accepted if its approximate likelihood (\autoref{eq:heatstress-likelihood}) is high enough, $\likelihood(\theta^*)>1-\epsilon_t$ (or more commonly, if $1-\likelihood(\theta^*) < \epsilon_t$), where $\epsilon_t>0$ is the \emph{acceptance threshold}, as higher values of $\epsilon_t$ allow more particles to be accepted. 
The acceptance threshold $\epsilon_t$ is chosen as the median of the $1-\likelihood(\theta)$ of the particles accepted in the previous iteration, $t-1$, and $\epsilon_0=0.01$. 
For each accepted particle $\theta_{i,t}$ a weight $w_t(\theta_{i,t})$ is assigned: for $t=0$, $w_0(\theta_{i,0})=1$, and for $t>0$, 
$w_t(\theta_{i,t}) = p(\theta_{i,t}) / \sum_{i=1}^{n_{t-1}}{w_{t-1}(\theta_{i,t-1}) K_t(\theta_{i,t}, \theta_{i,t-1})}$, where $p(\theta)$ is the prior density of $\theta$ and $K_t(\theta', \theta)$ is the probability of a perturbation from $\theta$ to $\theta'$.
$K_t(\theta' \mid \theta)$ is a multivariate normal distribution. % TODO construction of K_T from \theta_{i,t-1}

Acceptance is determined according to the approximate likelihood, which has a maximum value of 0.875. Thus, we terminated the inference when $\epsilon \le 0.13$ after six iterations, with $n_6=982$ accepted parameter vectors and effective sample size (ESS) 651 (\autoref{fig:convergence}). % TODO for multi-locus model we need to say something else
Running the inference algorithm with different initialization seeds and different number of simulations with produced similar outcomes (\autoref{fig:seeds}). %TODO for multi-locus model, write about them here also.

>>>>

\paragraph{Likelihood function.} 
Because our model, just like the Wright-Fisher model, is non-linear and stochastic, computing the distribution of fixation time $T(g)$ of genotype $g$ for use in the likelihood function is intractable (it is even hard to use a diffusion-equation approximation due to the model having multiple genotypes, rather than just two).
We overcome this problem by approximating the likelihood using simulations. We simulate 1,000 experiments per parameter vector $\theta = (\mu, \delta, s, b, c)$, resulting in a set of simulated observations $\tilde{\vec X} = \{\tilde{X}_i\}_{i=1}^{1000}$. We then compute the approximate likelihood,
\begin{equation}\begin{aligned}
\label{eq:heatstress-likelihood}
\likelihood(\theta) = &\ P^4(200 \le T(2n+1) \le 450) \cdot 
	\Big[1 - \\
	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)- \\
	&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big)+ \\
	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big) 
	\Big]\;,
\end{aligned}\end{equation}
where $!\{\ldots\}$ is the "logical not" operator, $P^4(\ldots)$ is the 4th power of $P(\ldots)$, and all probabilities $P_{\tilde{\vec X}}(\ldots)$ are approximated from the results of the simulations $\tilde{\vec X}$. For example, $P_{\tilde{\vec X}}\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)$ is approximated by taking simulations in which \anwt\; fixed before generation 450 but not before generation 200, and computing the fraction of such simulations in which \eumt\; did not fix by generation 1,700, and hence aneuploidy did not extinct before generation 1,700.
\autoref{fig:seeds} compares results with less and more simulated experiments, demonstrating that 1,000 simulations are likely enough.

@@

% TODO if we show results of multi-locus model, write how much replicates were used.

@@

The ABC-SMC algorithm employs sequential importance sampling over multiple iterations~\citep{Toni2009, Klinger2017, Syga2021}.
In iteration $t$ of the algorithm, a set of parameter vectors, $\{\theta_{i,t}\}_{i=1}^{n_t}$, also called \emph{particles}, are constructed in the following way.
A proposal particle, $\theta^*$, is sampled from a proposal distribution, and is either accepted or rejected, until $n_t$ particles are accepted.
The number of particles, $n_t$, is adapted at every iteration $t$ using the adaptive population strategy~\citep[\href{https://pyabc.readthedocs.io}{pyabc.readthedocs.io}]{Klinger2018}.
For $t=0$, the proposal particle is sampled from the prior distribution, $p(\theta)$.
For $t>0$, the proposal particle is sampled from the particles accepted in the previous iteration, $\{\theta_{i,t-1}\}_{i=1}^{n_{t-1}}$, each with a probability relative to its weight $W_{t-1}(\theta_{i,t-1})$ (see below). The proposal particle is then perturbed using a kernel perturbation kernel, $K_t(\theta^* \mid \theta)$ where $\theta$ is the sample from the previous iteration.
Then, a set of synthetic observations $\tilde{\vec X}^*$ is simulated, and the proposal particle $\theta^*$ is accepted if its approximate likelihood (\autoref{eq:heatstress-likelihood}) is high enough, $\likelihood(\theta^*)>1-\epsilon_t$ (or more commonly, if $1-\likelihood(\theta^*) < \epsilon_t$), where $\epsilon_t>0$ is the \emph{acceptance threshold}, as higher values of $\epsilon_t$ allow more particles to be accepted. 
The acceptance threshold $\epsilon_t$ is chosen as the median of the $1-\likelihood(\theta)$ of the particles accepted in the previous iteration, $t-1$, and $\epsilon_0=0.01$. 
For each accepted particle $\theta_{i,t}$ a weight $W_t(\theta_{i,t})$ is assigned: for $t=0$, $W_0(\theta_{i,0})=1$, and for $t>0$, 
$W_t(\theta_{i,t}) = p(\theta_{i,t}) / \sum_{i=1}^{n_{t-1}}{W_{t-1}(\theta_{i,t-1}) K_t(\theta_{i,t}, \theta_{i,t-1})}$, where $p(\theta)$ is the prior density of $\theta$ and $K_t(\theta', \theta)$ is the probability of a perturbation from $\theta$ to $\theta'$.
$K_t(\theta' \mid \theta)$ is a multivariate normal distribution. % TODO construction of K_T from \theta_{i,t-1}

Acceptance is determined according to the approximate likelihood~(\autoref{eq:heatstress-likelihood}), which has a maximum value of 0.875. Thus, we terminated the inference when $\epsilon \le 0.13$ after six iterations, with $n_6=982$ accepted parameter vectors and effective sample size ESS=651 (\autoref{fig:convergence}). % TODO for multi-locus model we need to say something else
Running the inference algorithm with different initialization seeds and less or more simulations for approximating the likelihood produced similar posterior distributions~(\autoref{fig:seeds}). %TODO for multi-locus model, write about them here also.
